---
title: "ML_class8"
author: "Phoebe He"
date: "2/6/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

** Clustering**

*K-means clustering*
```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```
```{r}
km <- kmeans(x,2,nstart=20)
km
```
Size of Clusters
```{r}
km$size
km$cluster
plot(x,col=km$cluster)
points(km$centers,col='blue',pch=18,cex=2)
```

*Hierarchical clustering*

```{r}
# First we need to calculate point (dis)similarity
#   as the Euclidean distance between observations
dist_matrix <- dist(x)
# The hclust() function returns a hierarchical
#  clustering model
hc <- hclust(d = dist_matrix)
# the print method is not so useful here
hc
```


```{r}
plot(hc)
abline(h=6,col='red')
abline(h=2.5,col='blue')
grp2 <- cutree(hc,h=6)
grp6 <- cutree(hc,h=2.5)
table(grp6)
```

```{r}
plot(x,col=grp2)
```
#using different linkage methods in R

```{r}
d <- dist_matrix

hc.complete <- hclust(d, method="complete")
plot(hc.complete)
hc.average  <- hclust(d, method="average")
plot(hc.average)
hc.single   <- hclust(d, method="single")
plot(hc.single)
```


made up overlapping data a bit more rell
```{r}
# Step 1. Generate some example data for clustering
a <- rbind(
  matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2),   # c1
  matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
  matrix(c(rnorm(50, mean = 1, sd = 0.3),           # c3
           rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(a) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(a)
# Step 3. Generate colors for known clusters
#         (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(a, col=col)
points(km$centers,col='blue',pch=18,cex=2)
```


```{r}
plot(a, col=col)
a.hc<- hclust(dist(a))
plot(a.hc)
grps3 <- cutree(a.hc,3)
plot(a,col=grps3)
```

*Principal Component Analysis (PCA)*
```{r}
mydata <- read.csv("https://tinyurl.com/expression-CSV",
row.names=1)
head(mydata)
```
*NOTE* prcomp() expects the samples to be rows and genes to be columns so we need to first transpose the matrix with the t() function!

```{r}
pca <- prcomp(t(mydata),scale=TRUE)
summary(pca)
```
square of the standard deviation is the variation
```{r}
dim(pca$x)
plot(pca$x[,1],pca$x[,2],
     xlab="PC1",ylab="PC2",
     ylim=c(-10,10))
```


```{r}
#percent variance is often more informative to look at
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
barplot(pca.var.per, main="Scree Plot",
        xlab="Principal Component", ylab="Percent Variation")

```

Make our PCA plot nice
```{r}
## A vector of colors for wt and ko samples
colvec <- as.factor( substr( colnames(mydata), 1, 2) )
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))
```



**PCA exercise**
```{r}
ukf<- read.csv('/Users/phoebehe/Desktop/BGGN213/class08 ML/UK_foods.csv')
dim(ukf)
# Note how the minus indexing works
rownames(ukf) <- ukf[,1]
ukf <- ukf[,-1]
head(ukf)
dim(ukf)

#assign row names from the first col of the data upon reading.
ukf <- read.csv('/Users/phoebehe/Desktop/BGGN213/class08 ML/UK_foods.csv',row.names=1)
```

```{r}
barplot(as.matrix(ukf), beside=T, col=rainbow(nrow(ukf)))
barplot(as.matrix(ukf), beside=F, col=rainbow(nrow(ukf)))
pairs(ukf, col=rainbow(10), pch=16)
```

```{r}
pca <- prcomp( t(ukf) )
summary(pca)
plot(pca$x[,1],pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], labels=colnames(ukf),col=c("orange","red","blue","green"))
```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
z <- summary(pca)
z$importance
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```


```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```

