---
title: "Lec6HW_Phoebe_He"
author: "Phoebe He"
date: "1/25/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Define function

The function below is able to take in an ID from PDB and plot out the corresponding protein's chian A's alpha carbon's Bfactor level.
The inputs are the PDB ID of the desired protein, in string.
To use the function, simply put in the PDB ID of your desired protein and run plot_pdb(your_input_PDB_ID)
The output is a plot showing the Bfactor trend levels of the alpha carbons on chain A of the input protin.
The plot also marks the regions for alpha helices and beta sheets.
```{r}
# Below defines function plot_pdb
library(bio3d)
plot_pdb <- function(x) {# The inputs are the PDB ID of the protein
  s1.chainA <- trim.pdb(read.pdb(x), chain="A", elety="CA")
  s1.b <- s1.chainA$atom$b
  plotb3(s1.b, 
         sse=s1.chainA,  # This line of code makrs the grey/black boxex on the top and bottom of the chart, which are indicators for regions of alpha helices(black) and beta sheets(grey)
         typ="l", 
         ylab="Bfactor")
}

# Test the function with three different inputs
plot_pdb("4AKE")
plot_pdb("1AKE") 
plot_pdb("1E4Y")
```


##Mock Bonus: plot overlays

This mock bonus uses points to overlay the Bfactor plots of chain A carbon items of protien B(red) on top of same plot of protein A(black).
However, only protein A's alpha helices and beta sheets are labeled in the graph.
```{r}
s3 <- read.pdb("1E4Y") # this is protein B
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")
s3.b <- s3.chainA$atom$b
plot_pdb("1AKE")# this is protein A
points(s3.b, typ='l',col="red")
```

